---
title: "Effect of PROTAC treatment on the immune landscape in young and old mice"
subtitle: "B Cell Subset Analysis and Differential Expression"
output:
  html_document:
    toc: true
    keep_md: false
    toc_float: false
    number_sections: true
    theme: default
    highlight: tango
    code_folding: none
pandoc_args:
  - "+RTS"
  - "-K16000m"
  - "-RTS"
date: "`r format(Sys.time(), '%m/%d/%y')`"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(Seurat)
library(dplyr)
library(ggplot2)
library(writexl)
library(downloadthis)
library(patchwork)
library(openxlsx)
library(stringr)
library(tibble)
library(DESeq2)
library(harmony)
library(loupeR)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.align="center")
set.seed(42)
```

```{r load_data, echo=FALSE, message=FALSE, warning=FALSE}
# Set parameters
DATASET <- "INTEGRATED"
BASE_PATH <- "/blue/zhangw/BCBSR_BIOINFORMATICS/hkates/TEMI/GE-8124_scrnaseq/02_R/"
ANALYSIS_DIR <- file.path(BASE_PATH, "integrated_analysis")
OUTPUT_DIR <- file.path(ANALYSIS_DIR, "analysis_outputs")
WEB_DIR = "/orange/cancercenter-dept/web/secure/zhangw/GE-8124/integrated"

# Load integrated Seurat object
obj <- readRDS(file.path(OUTPUT_DIR, "objects", "integrated_with_vdj.rds"))

# Set default assay to RNA
DefaultAssay(obj) <- "RNA"

# Join layers in Seurat v5 for compatibility
obj[["RNA"]] <- JoinLayers(obj[["RNA"]])

# Convert factor columns to character
metadata_cols <- colnames(obj@meta.data)
for(col in metadata_cols) {
  if(is.factor(obj@meta.data[[col]])) {
    obj@meta.data[[col]] <- as.character(obj@meta.data[[col]])
  }
}

# Create enhanced categorical metadata
obj$cluster_annotation <- paste0("Cluster_", obj$seurat_clusters)
obj$bcr_status <- ifelse(obj$has_bcr %in% TRUE, "BCR_positive", "BCR_negative")
obj$tcr_status <- ifelse(obj$has_tcr %in% TRUE, "TCR_positive", "TCR_negative")
obj$immune_type <- case_when(
  obj$has_bcr %in% TRUE ~ "B_cells",
  obj$has_tcr %in% TRUE ~ "T_cells",
  TRUE ~ "Other"
)

# Create analysis group labels
obj$analysis_group_clean <- case_when(
  obj$analysis_group == "AL_analysis" ~ "Aged_Lymph_node",
  obj$analysis_group == "AT_analysis" ~ "Aged_Tumor",
  obj$analysis_group == "YL_analysis" ~ "Young_Lymph_node",
  obj$analysis_group == "YT_analysis" ~ "Young_Tumor",
  TRUE ~ obj$analysis_group
)
obj$group_treatment <- paste(obj$analysis_group_clean, obj$treatment, sep = "_")

# Standardize tissue labels
obj$tissue <- case_when(
  grepl("^Lymph$", trimws(obj$tissue), ignore.case = TRUE) ~ "Lymph node",
  grepl("^Lymph node$", trimws(obj$tissue), ignore.case = TRUE) ~ "Lymph node",
  grepl("^Tumor$", trimws(obj$tissue), ignore.case = TRUE) ~ "Tumor",
  TRUE ~ as.character(obj$tissue)
)
```

# Study Overview

**Principal Investigator**: Dr. Wiezhou Zhang ([zhangw@ufl.edu](mailto:zhangw@ufl.edu))  
**Bioinformatics Analysis**: Dr. Heather Kates ([hkates@ufl.edu](mailto:hkates@ufl.edu))

## Experimental Design

This analysis examines changes in B cell population composition and gene expression related to PROTAC treatment (PZ15227) and to age comparing young and aged mice lymph node and tumor tissues.

**Sample Groups:**

- **AL**: Aged mice, Lymph node samples  
- **AT**: Aged mice, Tumor samples
- **YL**: Young mice, Lymph node samples
- **YT**: Young mice, Tumor samples

**Treatment Conditions:**

- Vehicle control (3 biological replicates per group)
- PZ15227 treatment (3 biological replicates per group)

## Data Generation and Processing

**Sequencing**: 10X Genomics single-cell RNA sequencing with cell hashing and VDJ-B/T sequencing was performed on all samples.

**Analysis Pipeline**: Raw sequencing data was processed through Cell Ranger Multi, followed by standard Seurat-based single-cell analysis including quality control, normalization, clustering, and cell type annotation. All sample groups were integrated using Harmony batch correction.

**Input Dataset**: The integrated dataset contains **`r ncol(obj)`** high-quality cells across all conditions, with **`r sum(obj$has_bcr, na.rm = TRUE)`** BCR-positive B cells identified through VDJ sequencing.

## Data Access

Complete datasets, intermediate objects, and reproducible code are available at: https://data.rc.ufl.edu/secure/cancercenter-dept/zhangw/GE-8124/  
Contact ([hkates@ufl.edu](mailto:hkates@ufl.edu)) for access credentials.

# B Cell Population Analysis

## Identifying B Cells Using VDJ Data

B cells were identified using BCR (B cell receptor) information from the VDJ-B sequencing data. This approach is more specific than using gene expression markers alone, as it captures cells that have successfully undergone V(D)J recombination.

```{r subset_bcells}
b_cells <- subset(obj, subset = bcr_status == "BCR_positive")

cat("Original dataset:", ncol(obj), "cells\n")
cat("B cells identified:", ncol(b_cells), "cells\n")
cat("B cell percentage:", round(ncol(b_cells)/ncol(obj)*100, 2), "%\n")
```

## Clustering B Cell Subpopulations

Because B cells represent a heterogeneous population with distinct functional states, this subset was re-analyzed to identify B cell sub-populations with better resolution than in the full dataset including re-normalization, dimensionality reduction, and clustering.

Multiple clustering resolutions (0.05-0.8) were tested, yielding 4-17 clusters. Resolution 0.1 was selected for producing 6 biologically interpretable clusters that capture major B cell functional states without over-clustering.

```{r recluster_bcells,echo=TRUE,message=FALSE}
b_cells <- NormalizeData(b_cells)
b_cells <- FindVariableFeatures(b_cells, selection.method = "vst", nfeatures = 2000)
b_cells <- ScaleData(b_cells)
b_cells <- RunPCA(b_cells, features = VariableFeatures(object = b_cells))
b_cells <- RunHarmony(b_cells, group.by.vars = "sample_id")
b_cells <- RunUMAP(b_cells, reduction = "harmony", dims = 1:30)
resolutions <- c(0.05, 0.1, 0.2, 0.3, 0.5, 0.8)
b_cells <- FindNeighbors(b_cells, reduction = "harmony", dims = 1:30)

# Store clustering results
clustering_results <- data.frame(
  Resolution = numeric(),
  Number_of_Clusters = numeric()
)

invisible(capture.output({
  for(res in resolutions) {
    b_cells <- FindClusters(b_cells, resolution = res)
    n_clusters <- length(unique(b_cells[[paste0("RNA_snn_res.", res)]][,1]))
    clustering_results <- rbind(clustering_results, 
                              data.frame(Resolution = res, 
                                       Number_of_Clusters = n_clusters))
  }
}))

Idents(b_cells) <- b_cells$RNA_snn_res.0.1
b_cells$bcell_clusters <- Idents(b_cells)

# Display results table
knitr::kable(clustering_results, 
             col.names = c("Resolution", "Number of Clusters"),
             caption = "B Cell Clustering Results at Different Resolutions")
```

## B Cell Cluster Visualization

The plots show the identified B cell subpopulations and how they are distributed across experimental conditions.

```{r visualize_clusters, fig.width=15, fig.height=12}
p1 <- DimPlot(b_cells, reduction = "umap", label = TRUE, pt.size = 0.5) +
  ggtitle("B Cell Clusters")

p2 <- DimPlot(b_cells, reduction = "umap", group.by = "age", pt.size = 0.5) +
  ggtitle("Age Groups")

p3 <- DimPlot(b_cells, reduction = "umap", group.by = "treatment", pt.size = 0.5) +
  ggtitle("Treatment Conditions")

p4 <- DimPlot(b_cells, reduction = "umap", group.by = "tissue", pt.size = 0.5) +
  ggtitle("Tissue Origin")

combined_plot <- (p1 | p2) / (p3 | p4)
print(combined_plot)

cluster_composition <- b_cells@meta.data %>%
  group_by(bcell_clusters, age, tissue, treatment) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(bcell_clusters) %>%
  mutate(percentage = count/sum(count)*100)

p5 <- ggplot(cluster_composition, aes(x = bcell_clusters, y = percentage, fill = interaction(age, treatment))) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~tissue) +
  theme_minimal() +
  labs(title = "B Cell Cluster Composition Across Experimental Conditions",
       x = "B Cell Cluster", y = "Percentage", fill = "Age × Treatment") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p5)
```

## B Cell Subpopulation Characterization

To identify distinct B cell functional states, differential expression analysis was performed to find cluster-specific marker genes. These markers were then used to annotate clusters into biologically meaningful B cell subtypes.

```{r find_markers, echo=FALSE, message=FALSE}
cluster_markers <- FindAllMarkers(b_cells, only.pos = TRUE, min.pct = 0.25,
                                  logfc.threshold = 0.25, test.use = "wilcox")
top_markers <- cluster_markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC) %>%
  arrange(cluster, desc(avg_log2FC))

cluster_markers %>% downloadthis::download_this(
      output_name = "BCells_ClusterMarkers",
      output_extension = ".xlsx",
      button_label = "Download B Cell Cluster Markers",
      button_type = "primary",
      has_icon = TRUE,
      icon = "fa fa-download"
    )

# Create summary table of top 3 markers per cluster
top_marker_summary <- top_markers %>%
  group_by(cluster) %>%
  slice_head(n = 3) %>%
  summarise(
    top_3_markers = paste(gene, collapse = ", "),
    avg_logfc_range = paste(round(range(avg_log2FC), 2), collapse = "-"),
    .groups = 'drop'
  )

knitr::kable(top_marker_summary, 
             col.names = c("Cluster", "Top 3 Markers", "Log2FC Range"),
             caption = "Top Differential Expression Markers by Cluster")
```

## Annotation Strategy: From Automated to Manual Curation

Initial attempts using automated reference-based annotation tools (SingleR, Azimuth) and predefined gene signature matching yielded inconsistent results with poor cluster resolution. Therefore, a manual curation approach was adopted using cluster-specific top markers and established B cell subtype signatures.

## Manual Marker Visualization

Key B cell subtype markers were visualized to assess cluster identity based on spatial expression patterns in the UMAP embedding.

```{manual viz}
# Cluster-specific top markers identified from differential expression
cluster_specific_markers <- c("Vpreb3", "Ifit3", "Cd80", "Themis", "Camk1d", "Tnfsf13")
available_cluster_markers <- cluster_specific_markers[cluster_specific_markers %in% rownames(b_cells)]

actual_marker_plots <- lapply(available_cluster_markers, function(marker) {
  FeaturePlot(b_cells, features = marker, pt.size = 0.5) +
    ggtitle(paste(marker, "expression")) +
    theme(plot.title = element_text(size = 10))
})

print(wrap_plots(actual_marker_plots, ncol = 3))

# Create annotation rationale table
annotation_rationale <- data.frame(
  Cluster = 0:5,
  Key_Marker = c("Vpreb3", "Ifit3", "Cd80", "Themis", "Camk1d", "Tnfsf13"),
  Marker_Function = c(
    "Pre-B cell receptor component",
    "Interferon-induced antiviral",
    "B cell activation/costimulation", 
    "T cell selection/activation",
    "Calcium signaling kinase",
    "B cell survival factor"
  ),
  Biological_Context = c(
    "B cell development",
    "Antiviral response", 
    "Classical activation",
    "Alternative activation",
    "Early development",
    "Interferon signaling"
  )
)

knitr::kable(annotation_rationale,
             caption = "Cluster Annotation Rationale Based on Key Markers")
```

### Final Manual Annotation

Based on the spatial expression patterns of cluster-specific markers and known B cell biology, clusters were manually annotated into functionally distinct subtypes.

```{r manual annotation}
cluster_annotations_manual <- data.frame(
  cluster = 0:5,
  annotation = c(
    "Transitional B cells",
    "Interferon-stimulated B cells", 
    "Activated B cells",
    "Alternative activated B cells",
    "Immature B cells",
    "Type I IFN B cells"
  ),
  n_cells = as.numeric(table(b_cells$bcell_clusters)),
  percentage = round(as.numeric(table(b_cells$bcell_clusters))/ncol(b_cells)*100, 1)
)

b_cells$manual_annotation <- cluster_annotations_manual$annotation[match(b_cells$bcell_clusters, 
                                                                        cluster_annotations_manual$cluster)]

knitr::kable(cluster_annotations_manual, 
             col.names = c("Cluster", "Manual Annotation", "Cell Count", "Percentage"),
             caption = "Final B Cell Cluster Annotations and Cell Distributions")
DimPlot(b_cells, group.by = "manual_annotation", label = TRUE, pt.size = 0.3, 
        label.size = 3, repel = TRUE) +
  ggtitle("Manually Annotated B Cell Subtypes") +
  theme(legend.position = "right",
        legend.text = element_text(size = 9))
```

## Download Loupe Browser File

A .cloupe file was created from the Seurat object of subsetted re-analyzed B-cells to enable interactive visualization and explorattion in the 10X loupe browser. [Download re-clustered annotated B-cells](https://data.rc.ufl.edu/secure/cancercenter-dept/zhangw/GE-8124/integrated/objects/BCells_Analysis.cloupe)

```{r create_loupe}
loupe_metadata <- b_cells@meta.data
loupe_metadata$Cluster <- paste0("Cluster_", loupe_metadata$bcell_clusters)
loupe_metadata$Age_Treatment <- paste(loupe_metadata$age, loupe_metadata$treatment, sep = "_")
loupe_metadata$Tissue_Age <- paste(loupe_metadata$tissue, loupe_metadata$age, sep = "_")

cluster_ann_df <- cluster_annotations_manual %>%
  mutate(cluster = as.character(cluster)) %>%
  select(cluster, annotation)  # Change this line

loupe_metadata <- loupe_metadata %>%
  rownames_to_column("cell_id") %>%
  left_join(cluster_ann_df, by = c("bcell_clusters" = "cluster")) %>%
  column_to_rownames("cell_id")

colnames(loupe_metadata) <- gsub(" ", "_", colnames(loupe_metadata))
colnames(loupe_metadata) <- gsub("-", "_", colnames(loupe_metadata))

b_cells@meta.data <- loupe_metadata

tryCatch({
  create_loupe_from_seurat(
    b_cells,
    output_name = "BCells_Analysis.cloupe",
    force = TRUE
  )
  loupe_created <- TRUE
}, error = function(e) {
  loupe_created <- FALSE
})
```

# Differential Expression Analysis

## Experimental Comparisons

Differential expression analysis was performed using the pseudobulk methodol, which aggregates cells within each sample before statistical testing. This approach properly accounts for biological replication by converting single-cell data to pseudobulk format by aggregating expression within each sample. This approach treats biological samples (not individual cells) as statistical units.and reduces false positives compared to cell-level testing.

**The six comparisons are:**

1. **Treatment effects in aged lymph node**: PZ15227 vs Vehicle in aged mice (lymph node)
2. **Treatment effects in aged tumor**: PZ15227 vs Vehicle in aged mice (tumor)  
3. **Treatment effects in young lymph node**: PZ15227 vs Vehicle in young mice (lymph node)
4. **Treatment effects in young tumor**: PZ15227 vs Vehicle in young mice (tumor)
5. **Age effects in lymph node**: Young vs Aged (vehicle-treated, lymph node)
6. **Age effects in tumor**: Young vs Aged (vehicle-treated, tumor)

```{r prepare_pseudobulk}
b_cells$pseudobulk_id <- paste(b_cells$bcell_clusters,
                              b_cells$group_treatment,
                              b_cells$sample_id,
                              sep = "_")

pseudobulk_data <- AggregateExpression(
  b_cells,
  assays = "RNA",
  slot = "counts",
  group.by = "pseudobulk_id",
  return.seurat = FALSE
)$RNA

pseudobulk_meta <- b_cells@meta.data %>%
  select(pseudobulk_id, bcell_clusters, group_treatment, sample_id) %>%
  distinct() %>%
  remove_rownames() %>%
  column_to_rownames("pseudobulk_id")

colnames(pseudobulk_data) <- gsub("^g", "", colnames(pseudobulk_data))
colnames(pseudobulk_data) <- gsub("-", "_", colnames(pseudobulk_data))

cat("Expression matrix dimensions:", dim(pseudobulk_data), "\n")
cat("Pseudobulk samples:", nrow(pseudobulk_meta), "\n")
```

## Statistical Analysis with DESeq2

DESeq2 was used for differential expression testing on pseudobulk data. This method properly models the negative binomial distribution of RNA-seq count data and controls for multiple testing across genes.

**Sample Size Requirements**: DESeq2 requires sufficient biological replicates to estimate gene-wise variance accurately. A minimum of 4 total samples are required per comparison, which allows for the missing YT_Vehicle_2 sample while maintaining statistical power.

```{r differential_expression}
run_deseq2_analysis <- function(counts, metadata, ref_group, test_group, cluster_num) {
  cluster_samples <- rownames(metadata)[metadata$bcell_clusters == cluster_num]
  if(length(cluster_samples) < 4) {
    return(NULL)
  }
  
  cluster_counts <- counts[, cluster_samples]
  cluster_meta <- metadata[cluster_samples, ]
  
  comparison_samples <- rownames(cluster_meta)[cluster_meta$group_treatment %in% c(ref_group, test_group)]
  if(length(comparison_samples) < 4) {
    return(NULL)
  }
  
  final_counts <- cluster_counts[, comparison_samples]
  final_meta <- cluster_meta[comparison_samples, ]
  final_meta$group_treatment <- factor(final_meta$group_treatment, levels = c(ref_group, test_group))
  
  keep <- rowSums(final_counts >= 10) >= 2
  final_counts <- final_counts[keep, ]
  
  dds <- DESeqDataSetFromMatrix(
    countData = final_counts,
    colData = final_meta,
    design = ~ group_treatment
  )
  
  dds <- DESeq(dds, quiet = TRUE)
  res <- results(dds, contrast = c("group_treatment", test_group, ref_group))
  
  res_df <- as.data.frame(res) %>%
    rownames_to_column("gene") %>%
    mutate(
      cluster = cluster_num,
      comparison = paste(test_group, "vs", ref_group),
      significant = padj < 0.05 & abs(log2FoldChange) > 0.5
    ) %>%
    arrange(padj)
  
  return(res_df)
}

comparisons <- list(
  "Aged_Lymphnode_PZ15227_vs_Vehicle" = c("Aged_Lymph_node_Vehicle", "Aged_Lymph_node_PZ15227"),
  "Aged_Tumor_PZ15227_vs_Vehicle" = c("Aged_Tumor_Vehicle", "Aged_Tumor_PZ15227"),
  "Young_Lymphnode_PZ15227_vs_Vehicle" = c("Young_Lymph_node_Vehicle", "Young_Lymph_node_PZ15227"),
  "Young_Tumor_PZ15227_vs_Vehicle" = c("Young_Tumor_Vehicle", "Young_Tumor_PZ15227"),
  "Young_vs_Aged_Lymphnode_Vehicle" = c("Aged_Lymph_node_Vehicle", "Young_Lymph_node_Vehicle"),
  "Young_vs_Aged_Tumor_Vehicle" = c("Aged_Tumor_Vehicle", "Young_Tumor_Vehicle")
)

all_results <- list()

for(comp_name in names(comparisons)) {
  ref_group <- comparisons[[comp_name]][1]
  test_group <- comparisons[[comp_name]][2]
  
  cluster_results <- list()
  for(cluster in unique(b_cells$bcell_clusters)) {
    result <- run_deseq2_analysis(
      pseudobulk_data,
      pseudobulk_meta,
      ref_group,
      test_group,
      cluster
    )
    if(!is.null(result)) {
      cluster_results[[paste0("Cluster_", cluster)]] <- result
    }
  }
  
  if(length(cluster_results) > 0) {
    all_results[[comp_name]] <- do.call(rbind, cluster_results)
  }
}

cat("Completed", length(all_results), "of", length(comparisons), "comparisons\n")
```

## Results Summary and Download

```{r create_and_download_results}
results_with_annotations <- list()

results_with_annotations[["Cluster_Annotations"]] <- cluster_annotations_manual

for(comp_name in names(all_results)) {
  sheet_name <- gsub("_", " ", comp_name)
  sheet_name <- str_to_title(sheet_name)
  
annotated_results <- all_results[[comp_name]] %>%
  left_join(cluster_annotations_manual %>% mutate(cluster = as.character(cluster)), 
            by = c("cluster" = "cluster")) %>%
  select(gene, cluster, log2FoldChange, padj, significant, comparison,
         annotation, n_cells, percentage, everything()) %>%
  arrange(cluster, padj)
  
  results_with_annotations[[sheet_name]] <- annotated_results
}

options(shiny.maxRequestSize = 100*1024^2)

tryCatch({
  results_with_annotations %>%
    downloadthis::download_this(
      output_name = "BCells_DE_Results_Complete",
      output_extension = ".xlsx",
      button_label = "Download B Cell DE Results",
      button_type = "primary",
      has_icon = TRUE,
      icon = "fa fa-download"
    )
}, error = function(e) {
  wb <- createWorkbook()
  for(sheet_name in names(results_with_annotations)) {
    addWorksheet(wb, sheet_name)
    writeData(wb, sheet_name, results_with_annotations[[sheet_name]])
  }
  saveWorkbook(wb, "BCells_DE_Results_Complete.xlsx", overwrite = TRUE)
})
```

## Analysis Summary

Overview of differential expression results across all comparisons.

```{r analysis_summary, fig.width=12, fig.height=6}
if(length(all_results) > 0) {
  summary_stats <- data.frame(
    Comparison = names(all_results),
    Total_Genes_Tested = sapply(all_results, function(x) length(unique(x$gene))),
    Significant_Genes = sapply(all_results, function(x) sum(x$significant, na.rm = TRUE)),
    Upregulated = sapply(all_results, function(x) sum(x$significant & x$log2FoldChange > 0, na.rm = TRUE)),
    Downregulated = sapply(all_results, function(x) sum(x$significant & x$log2FoldChange < 0, na.rm = TRUE)),
    Clusters_Analyzed = sapply(all_results, function(x) length(unique(x$cluster))),
    stringsAsFactors = FALSE
  )
  
  summary_stats$Comparison_Clean <- gsub("_", " ", summary_stats$Comparison)
  summary_stats$Comparison_Clean <- gsub("vs", "vs.", summary_stats$Comparison_Clean)
  
  print(summary_stats[, c("Comparison_Clean", "Significant_Genes", "Upregulated", "Downregulated", "Clusters_Analyzed")])
  
  p_summary <- summary_stats %>%
    ggplot(aes(x = reorder(Comparison_Clean, Significant_Genes), y = Significant_Genes)) +
    geom_col(fill = "steelblue", alpha = 0.8, width = 0.7) +
    coord_flip() +
    theme_minimal() +
    labs(title = "Significant Genes Identified Across Comparisons",
         subtitle = "Genes with padj < 0.05 and |log2FC| > 0.5",
         x = "Comparison", y = "Number of Significant Genes") +
    theme(
      axis.text.y = element_text(size = 10),
      plot.title = element_text(size = 14, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5)
    )
  
  print(p_summary)
  
  cluster_summary <- do.call(rbind, all_results) %>%
    group_by(cluster) %>%
    summarise(
      Total_Significant = sum(significant, na.rm = TRUE),
      Avg_LogFC = mean(abs(log2FoldChange[significant]), na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    arrange(desc(Total_Significant))
  
  cat("Significant genes by B cell cluster:\n")
  print(cluster_summary)
  
} else {
  cat("No differential expression results available.\n")
}
```
